---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
1. Unzip and load the source data file  
```{r loaddata}
## Unzip the source file
unzip("activity.zip")

## Load it in
sourceData <- read.csv("activity.csv")

## Examine the first few rows
head(sourceData)
```
2. No transformation of data required as it appears well structured with NA used for missing values.
  
## What is mean total number of steps taken per day?
1. Table showing daily total step count:
```{r dailytotalsteps, results='asis'}
## First create an aggregated(sum) dataset at interval level
dailyAggSteps <- setNames(
                                aggregate(sourceData$steps, list(sourceData$date), sum), 
                                c("Date", "Steps")
                        )

## Load the xtable library for neat display
library(xtable)

## Create the table of data
xt <- xtable(dailyAggSteps, pad=TRUE)

## And display it
print(xt, type="html")
```
  
  
2. Historgram of the above data showing the daily total step counts:  
```{r totalshistogram}
## Draw the histogram chart
with(dailyAggSteps, {
    par(las=2, mar=c(7,5,3,3), mgp=c(4,1,0))
    barplot(
        height=Steps,
        main="Total steps per day",
        xlab="Dates",
        cex.names=0.6,
        ylab="Step count",
        names.arg=Date,
        space=c(0)
    )
})
```


3. Calculate the mean and median total of steps per day (ignoring NA values):  
```{r meanmediancalcs}
## Calculate the mean from the daily aggregate dataset
dailyMeanSteps <- mean(dailyAggSteps$Steps, na.rm = TRUE)

## Calculate the median from the daily aggregate dataset
dailyMedianSteps <- median(dailyAggSteps$Steps, na.rm = TRUE)
```
  
The mean is **`r format(dailyMeanSteps, digits=6)`** and the median value is **`r dailyMedianSteps`**



## What is the average daily activity pattern?

1. A timeseries plot showing the average step count per interval:
```{r averagedailtyactivity}
## First create an aggregated(mean) dataset at interval level
intervalMeanSteps <- setNames(
                                aggregate(data=sourceData, steps~interval, FUN=mean, na.action=na.omit),
                                c("Interval", "MeanSteps")
                        )

## And then plot the result as a timeseries
with(intervalMeanSteps, {
    plot(
        x=Interval,
        y=MeanSteps,
        type="l",
        main="Average steps per interval across whole period",
        xlab="Interval",
        ylab="Mean step count"
        
    )
})
```
  
  
2. Find the 5-minute interval with the maximum number of steps
```{r maxstepinterval}
## Calculate the max value
intervalMax <- intervalMeanSteps[intervalMeanSteps$MeanSteps==max(intervalMeanSteps$MeanSteps),]
```
The maximum average step count occurs in interval **`r intervalMax$Interval`** to **`r intervalMax$Interval+5`** with an average **`r format(intervalMax$MeanSteps, digits=5)`** steps.


## Imputing missing values

1. Count of the number of rows with missing values:
```{r missingvalues}
## Do the calculation
countNA <- nrow(subset(sourceData, is.na(sourceData$steps)))
```
  
There are **`r countNA`** rows with missing values.
  
2. We'll use the previously calculated 5-minute interval average step counts to replace NA values.
3. Create the new dataset with the missing data filled in:
```{r imputemissingvalues}
##First get the original step counts into a dataframe
stepCountValues <- data.frame(sourceData$steps)

## Use tapply to find and replace NA values with corresponding averages (rounded up to full steps)
stepCountValues[is.na(stepCountValues),] <- 
        ceiling(tapply(X=sourceData$steps, INDEX=sourceData$interval, FUN=mean, na.rm=TRUE))

## Bind the new and original data into an imputed dataset
imputedData <- setNames(
                                cbind(stepCountValues, sourceData[,2:3]),
                                c("Steps", "Date", "Interval")
                        )

## Examine the resulting dataset
head(imputedData)
```

4. A historgram showing the new daily total step counts: 
```{r newdailymeansteps}
## Firast create an aggregated(sum) dataset at day level for imputed dataset
newDailyAggSteps <- setNames(
                                aggregate(imputedData$Steps, list(imputedData$Date), sum), 
                                c("Date", "Steps")
                        )
## Calculate the mean and median total of steps per day (missing values imputed)
newDailyMeanSteps <- mean(newDailyAggSteps$Steps, na.rm = TRUE)
newDailyMedianSteps <- median(newDailyAggSteps$Steps, na.rm = TRUE)

## Now draw a historgram showing the new daily total step counts 
with(newDailyAggSteps, {
    par(las=2, mar=c(7,5,3,3), mgp=c(4,1,0))
    barplot(
        height=Steps,
        main="Total steps per day",
        xlab="Dates",
        cex.names=0.6,
        ylab="Step count",
        names.arg=Date,
        space=c(0)
    )
})
```

Replacing missing values with the interval average caused both mean and median values to increase with the Mean going from **`r format(dailyMeanSteps, digits=6)`** to **`r format(newDailyMeanSteps, digits=6)`** and the Median going from **`r format(dailyMedianSteps, digits=5)`** to **`r format(newDailyMedianSteps, digits=5)`**.

## Are there differences in activity patterns between weekdays and weekends?
  
1. Classify the rows by day type; weekday or weekend and calculate average steps by that classification.
```{r classifybydaytype}
## Use tapply and custom func to classify dates as weekday or weekend and add to dataframe
dateDayClass <- data.frame(sapply(X = imputedData$Date, FUN = 
                        function(day) {
                                if (weekdays(as.Date(day)) %in% 
                                    c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {
                                        day <- "weekday"
                                } 
                                else {
                                        day <- "weekend"
                                }
                        }
                ))
## Bind result back to target dataset and name the cols
newDataWithDayClass <- setNames(cbind(imputedData, dateDayClass),
                                c("Steps", "Date", "Interval", "DayClass"))

## Calculate the mean number of steps by interval by day classification:
dayClassMeanIntSteps <- aggregate(
    data=newDataWithDayClass,
    Steps ~ DayClass + Interval,
    FUN=mean
)

## Examine the result
head(dayClassMeanIntSteps)
```

2. Timeseries panel plot showing the average step count per interval for both weekday and weekend day classes:
```{r panelplotbydayclass}
## Load the lattice package
library("lattice")
## Create the panel plot
xyplot(
    type="l",
    data=dayClassMeanIntSteps,
    Steps ~ Interval | DayClass,
    xlab="Interval",
    ylab="Mean step count",
    layout=c(1,2)
)
```